---
title: "Predicting Loan Interest Rates: Regression Trees vs Linear Models"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5.5)
```

## What This Project Is About

When someone applies for a personal loan on Lending Club, the platform assigns them an interest rate. That rate is not random. It reflects the platform's assessment of how risky the borrower is. A borrower with a high credit score, stable income, and a short loan term will get a lower rate. Someone with past delinquencies, high debt relative to income, and a longer loan term will pay more.

This project builds three models that each try to predict what interest rate a borrower will receive, given information about their financial background. The three models are a regression tree, a linear regression with best subset selection, and a LASSO regression. The goal is to compare how well each method predicts interest rates on borrowers the model has never seen, and to understand what each model is actually learning about who pays what.

The dataset comes from Lending Club, one of the largest peer-to-peer lending platforms in the US. It contains 15,000 loan records with 15 features per borrower and a target variable: the assigned interest rate.

---

## The Data

```{r load-data}
lend <- read.csv("Lending Club.csv")
cat("Rows:", nrow(lend), "\n")
cat("Columns:", ncol(lend), "\n")
cat("\nColumn names:\n")
print(names(lend))
```

```{r feature-table}
feature_df <- data.frame(
  Feature = c("Loan_amt", "Loan_term", "emp_length", "Home",
              "Income_source_verified", "Income_verified", "Income_thou",
              "Debt_income", "delinq_2yrs", "Credit_history_length",
              "FICO", "open_acc", "Derogatory_recs",
              "Revol_balance", "Revol_util", "Loan_rate"),
  Type = c("Numeric", "Numeric (36 or 60)", "Categorical", "Categorical",
           "Binary", "Binary", "Numeric",
           "Numeric", "Numeric", "Numeric",
           "Numeric", "Numeric", "Numeric",
           "Numeric", "Numeric", "Numeric (target)"),
  Description = c(
    "Loan amount requested in dollars",
    "Loan length in months (36 or 60)",
    "How many years the borrower has been employed",
    "Whether the borrower rents, owns, or has a mortgage",
    "Whether the income source was verified by Lending Club",
    "Whether the income amount was verified",
    "Annual income in thousands of dollars",
    "Debt-to-income ratio: monthly debt payments divided by monthly income",
    "Number of times the borrower was 30+ days late on a payment in the last 2 years",
    "How many years since the borrower opened their first credit account",
    "FICO credit score (300 to 850, higher is better)",
    "Number of currently open credit accounts",
    "Number of derogatory public records (bankruptcies, tax liens, etc.)",
    "Total revolving credit balance in dollars",
    "Revolving credit utilization: how much of available revolving credit is being used",
    "Interest rate assigned by Lending Club (what we are predicting)"
  )
)
knitr::kable(feature_df, caption = "Dataset features and descriptions")
```

```{r prep}
lend$emp_length <- factor(lend$emp_length)
lend$Home       <- factor(lend$Home)
cat("Employment length categories:", paste(levels(lend$emp_length), collapse = ", "), "\n")
cat("Home ownership categories:   ", paste(levels(lend$Home), collapse = ", "), "\n")
```

### Distribution of Interest Rates

```{r rate-dist}
library(ggplot2)

ggplot(lend, aes(x = Loan_rate)) +
  geom_histogram(binwidth = 0.5, fill = "#2c7bb6", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(lend$Loan_rate), linetype = "dashed",
             color = "#E07B39", linewidth = 1) +
  annotate("text", x = mean(lend$Loan_rate) + 0.5, y = 500,
           label = paste0("Mean: ", round(mean(lend$Loan_rate), 2), "%"),
           color = "#E07B39", hjust = 0, size = 3.8) +
  labs(title = "Distribution of Loan Interest Rates",
       subtitle = "15,000 Lending Club borrowers",
       x = "Interest Rate (%)", y = "Number of Borrowers") +
  theme_minimal() +
  theme(plot.subtitle = element_text(color = "grey40", size = 10))
```

The distribution is right-skewed. Most borrowers receive rates between 8% and 18%, but a meaningful tail extends to 26%. The average rate is about 13.8%. The spread is wide: a borrower at the low end pays roughly 6%, while someone at the high end pays over 26%. That 20 percentage point spread represents thousands of dollars in additional interest cost over the life of a loan.

### Loan Rate by Key Variables

```{r eda-charts, fig.width=10, fig.height=5}
library(gridExtra)

p1 <- ggplot(lend, aes(x = FICO, y = Loan_rate)) +
  geom_point(alpha = 0.08, color = "#2c7bb6", size = 0.8) +
  geom_smooth(method = "lm", color = "#E07B39", se = FALSE, linewidth = 1) +
  labs(title = "FICO Score vs Interest Rate",
       x = "FICO Score", y = "Interest Rate (%)") +
  theme_minimal()

p2 <- ggplot(lend, aes(x = factor(Loan_term), y = Loan_rate, fill = factor(Loan_term))) +
  geom_boxplot(alpha = 0.8, outlier.size = 0.5) +
  scale_fill_manual(values = c("36" = "#2c7bb6", "60" = "#E07B39")) +
  labs(title = "Loan Term vs Interest Rate",
       x = "Loan Term (months)", y = "Interest Rate (%)") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p1, p2, ncol = 2)
```

Two patterns stand out immediately. First, FICO score and interest rate have a strong negative relationship: higher credit scores consistently receive lower rates. This is the core logic of credit risk pricing. Second, 60-month loans carry significantly higher rates than 36-month loans. Longer loans expose the lender to more uncertainty about whether the borrower will remain able to repay.

```{r eda-charts2, fig.width=10, fig.height=5}
p3 <- ggplot(lend, aes(x = Debt_income, y = Loan_rate)) +
  geom_point(alpha = 0.08, color = "#3A9E82", size = 0.8) +
  geom_smooth(method = "lm", color = "#E07B39", se = FALSE, linewidth = 1) +
  labs(title = "Debt-to-Income Ratio vs Interest Rate",
       x = "Debt-to-Income Ratio (%)", y = "Interest Rate (%)") +
  theme_minimal()

p4 <- ggplot(lend, aes(x = Home, y = Loan_rate, fill = Home)) +
  geom_boxplot(alpha = 0.8, outlier.size = 0.5) +
  scale_fill_manual(values = c("MORTGAGE" = "#2c7bb6", "OWN" = "#3A9E82", "RENT" = "#E07B39")) +
  labs(title = "Home Ownership vs Interest Rate",
       x = "Home Ownership Status", y = "Interest Rate (%)") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p3, p4, ncol = 2)
```

Debt-to-income ratio has a mild positive relationship with interest rate: borrowers who are already carrying more debt relative to their income tend to receive slightly higher rates. Home ownership shows a clear pattern: renters pay the highest rates on average, homeowners with a mortgage are in the middle, and outright homeowners get the best rates, likely because outright ownership signals greater financial stability.

---

## Train and Validation Split

The data is split 80/20: 12,000 loans for training and 3,000 for validation. All three models are trained on the same training set and evaluated on the same validation set, so the RMS errors are directly comparable.

```{r split}
set.seed(1)
trainindex <- sample(15000, 12000)
train <- lend[trainindex, ]
valid <- lend[-trainindex, ]

cat("Training set:", nrow(train), "loans\n")
cat("Validation set:", nrow(valid), "loans\n")
cat("\nTraining set rate stats:\n")
cat("  Mean:", round(mean(train$Loan_rate), 2), "%\n")
cat("  SD:  ", round(sd(train$Loan_rate), 2), "%\n")
cat("\nValidation set rate stats:\n")
cat("  Mean:", round(mean(valid$Loan_rate), 2), "%\n")
cat("  SD:  ", round(sd(valid$Loan_rate), 2), "%\n")
```

The mean and standard deviation of interest rates are nearly identical in both sets, confirming the random split produced representative samples.

---

## Model 1: Regression Tree

A regression tree works by repeatedly splitting the data into subgroups based on feature thresholds. At each split, it picks the feature and threshold that most reduces the prediction error within the resulting subgroups. Once splitting stops, each terminal node (leaf) makes a single prediction: the average interest rate of all training loans that fell into that leaf.

The appeal of a regression tree is interpretability. You can follow the path from the root to any leaf and read exactly why a borrower received a particular predicted rate. The trade-off is that trees can overfit if grown too deep, and they often miss smooth continuous relationships that a linear model handles naturally.

```{r tree-model}
library(tree)

lendtree <- tree(Loan_rate ~ ., data = train)
summary(lendtree)
```

```{r tree-plot, fig.width=11, fig.height=7}
plot(lendtree, type = "uniform")
text(lendtree, pretty = 0, cex = 0.8, col = "#2c7bb6")
title(main = "Full Regression Tree: Predicting Loan Interest Rate", cex.main = 1.1)
```

The full tree uses only two variables out of the 15 available: loan term and FICO score. This is a strong signal. Even though the dataset has income, employment history, debt levels, delinquencies, and many other features, the tree finds that just those two variables explain most of the variation in interest rates. FICO score captures creditworthiness comprehensively, and loan term directly affects the risk premium the platform charges.

### Cross-Validation to Find Optimal Tree Size

```{r cv-tree, fig.width=8, fig.height=5}
cv <- cv.tree(lendtree)

cv_df <- data.frame(size = cv$size, rms = sqrt(cv$dev / nrow(train)))

ggplot(cv_df, aes(x = size, y = rms)) +
  geom_line(color = "#2c7bb6", linewidth = 1) +
  geom_point(color = "#2c7bb6", size = 3) +
  geom_vline(xintercept = 6, linetype = "dashed", color = "#E07B39", linewidth = 1) +
  annotate("text", x = 6.15, y = max(cv_df$rms) * 0.99,
           label = "Optimal: 6 nodes", color = "#E07B39", hjust = 0, size = 3.8) +
  labs(title = "Cross-Validation RMS Error by Tree Size",
       subtitle = "Lowest error is achieved at the full tree size (6 nodes)",
       x = "Number of Terminal Nodes", y = "CV RMS Error (%)") +
  theme_minimal() +
  theme(plot.subtitle = element_text(color = "grey40", size = 10))
```

Cross-validation tests each possible tree size and measures prediction error on held-out portions of the training data. Here, the full 6-node tree achieves the lowest cross-validation error. Pruning would increase error by removing splits that are genuinely useful. So the full tree is also the final tree.

```{r pruned-tree}
lendtree2 <- prune.tree(lendtree, best = 6)
```

```{r pruned-tree-plot, fig.width=11, fig.height=7}
library(rpart)
library(rpart.plot)

lendtree3 <- rpart(Loan_rate ~ ., data = train)

rpart.plot(lendtree3,
           type = 4, extra = 101,
           box.palette = "Blues",
           shadow.col = "grey80",
           nn = FALSE,
           main = "Regression Tree: Predicting Loan Interest Rate")
```

This cleaner version of the tree shows the same structure with predicted rates and observation counts at each leaf. Reading from the top:

- The first split is on loan term. Loans with 36-month terms go left, 60-month terms go right. 60-month loans already carry a rate premium just from their length.
- Within each side, FICO score determines where you end up. Higher FICO scores push borrowers into lower-rate leaves, lower FICO scores into higher-rate leaves.
- The lowest predicted rate (~8.3%) goes to 36-month borrowers with FICO above 745. The highest (~19.5%) goes to 60-month borrowers with FICO below 660.

### Tree Predictions vs Actual Rates

```{r tree-pred}
pred_tree <- predict(lendtree2, newdata = valid)
validRMS_tree <- sqrt(sum((valid$Loan_rate - pred_tree)^2) / nrow(valid))
cat("Regression tree RMS error on validation set:", round(validRMS_tree, 4), "%\n")
```

```{r tree-residuals, fig.width=9, fig.height=5}
resid_df <- data.frame(
  Actual    = valid$Loan_rate,
  Predicted = pred_tree,
  Residual  = valid$Loan_rate - pred_tree
)

p_pred <- ggplot(resid_df, aes(x = Predicted, y = Actual)) +
  geom_point(alpha = 0.15, color = "#2c7bb6", size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "#E07B39", linewidth = 1) +
  labs(title = "Predicted vs Actual: Regression Tree",
       x = "Predicted Rate (%)", y = "Actual Rate (%)") +
  theme_minimal()

p_resid <- ggplot(resid_df, aes(x = Predicted, y = Residual)) +
  geom_point(alpha = 0.15, color = "#2c7bb6", size = 0.8) +
  geom_hline(yintercept = 0, color = "#E07B39", linewidth = 1) +
  labs(title = "Residuals: Regression Tree",
       x = "Predicted Rate (%)", y = "Residual (Actual minus Predicted)") +
  theme_minimal()

grid.arrange(p_pred, p_resid, ncol = 2)
```

The predicted vs actual chart shows the limitation of the regression tree clearly. Because the tree only has 6 leaf nodes, it can only predict 6 distinct values. You can see the predicted values clustering in vertical bands. The tree is partitioning borrowers into buckets but cannot make a smooth, continuous prediction the way a linear model can. Borrowers within the same leaf all receive the same predicted rate regardless of how different their individual profiles are.

---

## Model 2: Linear Regression with Best Subset Selection

Best subset selection tries every possible combination of features and finds the one that produces the best model according to a criterion. Here we use adjusted R-squared, which rewards models that explain more variance while penalizing unnecessary complexity.

The reason to do this rather than simply including all features is that not every feature improves predictions. Some are redundant or add noise. Subset selection finds the most useful combination.

```{r subset-selection, fig.width=10, fig.height=5.5}
library(leaps)

subset_model <- regsubsets(Loan_rate ~ ., data = train, nvmax = 20)
subset_summary <- summary(subset_model)

adj_r2_df <- data.frame(
  num_vars = 1:length(subset_summary$adjr2),
  adj_r2   = subset_summary$adjr2
)
best_n <- which.max(subset_summary$adjr2)

ggplot(adj_r2_df, aes(x = num_vars, y = adj_r2)) +
  geom_line(color = "#2c7bb6", linewidth = 1) +
  geom_point(color = "#2c7bb6", size = 3) +
  geom_vline(xintercept = best_n, linetype = "dashed",
             color = "#E07B39", linewidth = 1) +
  annotate("text", x = best_n + 0.3, y = min(adj_r2_df$adj_r2) + 0.01,
           label = paste0("Best: ", best_n, " variables"),
           color = "#E07B39", hjust = 0, size = 3.8) +
  scale_x_continuous(breaks = 1:max(adj_r2_df$num_vars)) +
  labs(title = "Best Subset Selection: Adjusted R-squared by Model Size",
       subtitle = "Higher adjusted R-squared means better model fit, penalized for complexity",
       x = "Number of Variables in Model", y = "Adjusted R-squared") +
  theme_minimal() +
  theme(plot.subtitle = element_text(color = "grey40", size = 10))
```

```{r best-vars}
coef_best <- coef(subset_model, best_n)
cat("Best model uses", best_n, "variables:\n")
print(round(coef_best, 4))
```

```{r lm-predict}
# Build the best variable names and fit final linear model
best_vars <- names(coef_best)[-1]

# Handle factor dummy variable names back to original column names
clean_vars <- unique(gsub("MORTGAGE|OWN|RENT|< 1 year|1 year|2 years|3 years|4 years|5 years|6 years|7 years|8 years|9 years|10\\+ years|n/a", "", best_vars))
clean_vars <- unique(trimws(clean_vars))
clean_vars <- clean_vars[clean_vars != ""]

# Use the full formula approach: fit on all vars, subset selection just tells us which matter
# Build formula from best_vars
build_formula <- function(coef_names, response = "Loan_rate") {
  vars <- coef_names[coef_names != "(Intercept)"]
  # strip factor level suffixes to get original column names
  orig <- gsub("(Home|emp_length)(.*)", "\\1", vars)
  orig <- unique(orig)
  as.formula(paste(response, "~", paste(orig, collapse = " + ")))
}

lm_formula <- build_formula(names(coef_best))
cat("Formula used:\n")
print(lm_formula)

lm_model <- lm(lm_formula, data = train)
summary(lm_model)
```

```{r lm-coef-plot, fig.width=9, fig.height=5.5}
lm_coef <- as.data.frame(summary(lm_model)$coefficients)
lm_coef$Variable <- rownames(lm_coef)
lm_coef <- lm_coef[lm_coef$Variable != "(Intercept)", ]
lm_coef$Significant <- lm_coef$`Pr(>|t|)` < 0.05

readable <- c(
  "Loan_term"              = "Loan term (months)",
  "FICO"                   = "FICO credit score",
  "Debt_income"            = "Debt-to-income ratio",
  "Income_thou"            = "Annual income (thousands)",
  "Loan_amt"               = "Loan amount",
  "delinq_2yrs"            = "Delinquencies (last 2 yrs)",
  "Revol_util"             = "Revolving credit utilization",
  "open_acc"               = "Open credit accounts",
  "Credit_history_length"  = "Credit history length",
  "Derogatory_recs"        = "Derogatory public records",
  "Revol_balance"          = "Revolving credit balance",
  "Income_source_verified" = "Income source verified",
  "Income_verified"        = "Income verified",
  "HomeOWN"                = "Home: owns outright",
  "HomeRENT"               = "Home: renting",
  "HomeMORTGAGE"           = "Home: mortgage"
)

lm_coef$Label <- ifelse(lm_coef$Variable %in% names(readable),
                        readable[lm_coef$Variable],
                        lm_coef$Variable)

ggplot(lm_coef, aes(x = reorder(Label, Estimate), y = Estimate, fill = Significant)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "grey75", "TRUE" = "#2c7bb6")) +
  labs(title = "Linear Regression Coefficients",
       subtitle = "Blue bars are statistically significant (p < 0.05). Positive values push the rate higher.",
       x = NULL, y = "Coefficient Estimate",
       fill = "Significant") +
  theme_minimal() +
  theme(plot.subtitle = element_text(color = "grey40", size = 9))
```

```{r lm-rms}
pred_lm <- predict(lm_model, newdata = valid)
validRMS_lm <- sqrt(sum((valid$Loan_rate - pred_lm)^2) / nrow(valid))
cat("Linear regression RMS error on validation set:", round(validRMS_lm, 4), "%\n")
```

```{r lm-residuals, fig.width=9, fig.height=5}
resid_lm_df <- data.frame(
  Actual    = valid$Loan_rate,
  Predicted = pred_lm,
  Residual  = valid$Loan_rate - pred_lm
)

p_lm1 <- ggplot(resid_lm_df, aes(x = Predicted, y = Actual)) +
  geom_point(alpha = 0.15, color = "#3A9E82", size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "#E07B39", linewidth = 1) +
  labs(title = "Predicted vs Actual: Linear Regression",
       x = "Predicted Rate (%)", y = "Actual Rate (%)") +
  theme_minimal()

p_lm2 <- ggplot(resid_lm_df, aes(x = Predicted, y = Residual)) +
  geom_point(alpha = 0.15, color = "#3A9E82", size = 0.8) +
  geom_hline(yintercept = 0, color = "#E07B39", linewidth = 1) +
  labs(title = "Residuals: Linear Regression",
       x = "Predicted Rate (%)", y = "Residual") +
  theme_minimal()

grid.arrange(p_lm1, p_lm2, ncol = 2)
```

The linear regression predicted vs actual chart looks much tighter than the regression tree. Instead of 6 fixed values, the linear model makes a continuous prediction for each borrower. The predictions spread across the full range of actual rates. The residuals are centered around zero with no obvious pattern, which is what you want to see.

---

## Model 3: LASSO Regression

LASSO (Least Absolute Shrinkage and Selection Operator) is a form of linear regression that adds a penalty for including too many variables. As the penalty strength (lambda) increases, LASSO shrinks some coefficients toward zero and eventually sets them exactly to zero, effectively removing those variables from the model. This is called regularization.

The practical benefit is that LASSO automatically handles feature selection. You give it all the variables and it figures out which ones matter. It also tends to generalize better than ordinary linear regression when some features are noisy or correlated with each other.

```{r lasso, fig.width=9, fig.height=5}
library(glmnet)

x_train <- model.matrix(Loan_rate ~ ., data = train)[, -1]
y_train <- train$Loan_rate
x_valid <- model.matrix(Loan_rate ~ ., data = valid)[, -1]
y_valid  <- valid$Loan_rate

cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1)

lasso_cv_df <- data.frame(
  log_lambda = log(cv_lasso$lambda),
  mse        = cv_lasso$cvm,
  upper      = cv_lasso$cvup,
  lower      = cv_lasso$cvlo
)

ggplot(lasso_cv_df, aes(x = log_lambda, y = mse)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#2c7bb6", alpha = 0.15) +
  geom_line(color = "#2c7bb6", linewidth = 1) +
  geom_vline(xintercept = log(cv_lasso$lambda.min), linetype = "dashed",
             color = "#E07B39", linewidth = 1) +
  annotate("text",
           x = log(cv_lasso$lambda.min) + 0.15,
           y = max(lasso_cv_df$mse) * 0.98,
           label = paste0("Best lambda = ", round(cv_lasso$lambda.min, 4)),
           color = "#E07B39", hjust = 0, size = 3.5) +
  labs(title = "LASSO Cross-Validation: MSE by Lambda",
       subtitle = "Shaded band shows standard error. Dashed line marks the lambda with lowest error.",
       x = "Log(Lambda)", y = "Mean Squared Error") +
  theme_minimal() +
  theme(plot.subtitle = element_text(color = "grey40", size = 10))
```

As lambda increases (moving right on the chart), the model becomes more penalized and error rises. The dashed line marks the lambda value that minimizes cross-validated MSE. That is the lambda used for final predictions.

```{r lasso-coefs}
best_lambda <- cv_lasso$lambda.min
lasso_coefs <- coef(cv_lasso, s = best_lambda)
coef_df <- data.frame(
  Variable    = rownames(lasso_coefs),
  Coefficient = as.vector(lasso_coefs)
)
coef_df <- coef_df[coef_df$Variable != "(Intercept)" & coef_df$Coefficient != 0, ]
coef_df <- coef_df[order(abs(coef_df$Coefficient), decreasing = TRUE), ]

coef_df$Coefficient <- round(coef_df$Coefficient, 5)
cat("LASSO non-zero coefficients at best lambda:\n")
knitr::kable(coef_df,
             caption = paste0("LASSO coefficients (lambda = ", round(best_lambda, 4), ")"),
             row.names = FALSE)
```

```{r lasso-coef-plot, fig.width=9, fig.height=5.5}
coef_df$Direction <- ifelse(coef_df$Coefficient > 0, "Increases rate", "Decreases rate")

ggplot(coef_df, aes(x = reorder(Variable, Coefficient), y = Coefficient, fill = Direction)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("Increases rate" = "#E07B39", "Decreases rate" = "#2c7bb6")) +
  labs(title = "LASSO Regression Coefficients at Best Lambda",
       subtitle = "Only non-zero coefficients shown. Variables set to zero were removed by LASSO.",
       x = NULL, y = "Coefficient", fill = NULL) +
  theme_minimal() +
  theme(plot.subtitle = element_text(color = "grey40", size = 9),
        legend.position = "bottom")
```

```{r lasso-rms}
pred_lasso <- predict(cv_lasso, s = best_lambda, newx = x_valid)
validRMS_lasso <- sqrt(sum((y_valid - pred_lasso)^2) / nrow(valid))
cat("LASSO RMS error on validation set:", round(validRMS_lasso, 4), "%\n")
```

```{r lasso-residuals, fig.width=9, fig.height=5}
resid_lasso_df <- data.frame(
  Actual    = y_valid,
  Predicted = as.vector(pred_lasso),
  Residual  = y_valid - as.vector(pred_lasso)
)

p_las1 <- ggplot(resid_lasso_df, aes(x = Predicted, y = Actual)) +
  geom_point(alpha = 0.15, color = "#E07B39", size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "#2c7bb6", linewidth = 1) +
  labs(title = "Predicted vs Actual: LASSO",
       x = "Predicted Rate (%)", y = "Actual Rate (%)") +
  theme_minimal()

p_las2 <- ggplot(resid_lasso_df, aes(x = Predicted, y = Residual)) +
  geom_point(alpha = 0.15, color = "#E07B39", size = 0.8) +
  geom_hline(yintercept = 0, color = "#2c7bb6", linewidth = 1) +
  labs(title = "Residuals: LASSO",
       x = "Predicted Rate (%)", y = "Residual") +
  theme_minimal()

grid.arrange(p_las1, p_las2, ncol = 2)
```

---

## Model Comparison

```{r comparison, fig.width=9, fig.height=5}
results <- data.frame(
  Model = c("Regression Tree", "Linear Regression (Best Subset)", "LASSO Regression"),
  RMS_Error = c(round(validRMS_tree, 4),
                round(validRMS_lm,   4),
                round(validRMS_lasso, 4)),
  Features_Used = c("2 (Loan_term, FICO)",
                    paste(best_n, "selected by subset search"),
                    paste(nrow(coef_df), "non-zero after penalization"))
)
knitr::kable(results,
             caption = "Model performance on validation set (lower RMS error is better)",
             col.names = c("Model", "RMS Error (%)", "Features Used"))
```

```{r comparison-chart, fig.width=8, fig.height=4.5}
ggplot(results, aes(x = reorder(Model, RMS_Error), y = RMS_Error, fill = Model)) +
  geom_bar(stat = "identity", width = 0.55) +
  geom_text(aes(label = paste0(RMS_Error, "%")), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c(
    "Regression Tree"                  = "#2c7bb6",
    "Linear Regression (Best Subset)"  = "#3A9E82",
    "LASSO Regression"                 = "#E07B39"
  )) +
  scale_y_continuous(limits = c(0, max(results$RMS_Error) * 1.15)) +
  labs(title = "Validation Set RMS Error by Model",
       subtitle = "Lower is better",
       x = NULL, y = "RMS Error (%)") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.subtitle = element_text(color = "grey40", size = 10))
```

```{r error-dist, fig.width=9, fig.height=5}
all_resid <- data.frame(
  Residual = c(resid_df$Residual, resid_lm_df$Residual, resid_lasso_df$Residual),
  Model    = rep(c("Regression Tree", "Linear Regression", "LASSO"),
                 each = nrow(valid))
)

ggplot(all_resid, aes(x = Residual, fill = Model, color = Model)) +
  geom_density(alpha = 0.25, linewidth = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values  = c("Regression Tree" = "#2c7bb6",
                                "Linear Regression" = "#3A9E82",
                                "LASSO" = "#E07B39")) +
  scale_color_manual(values = c("Regression Tree" = "#2c7bb6",
                                "Linear Regression" = "#3A9E82",
                                "LASSO" = "#E07B39")) +
  labs(title = "Distribution of Prediction Errors by Model",
       subtitle = "Tighter distribution centered at zero means more accurate predictions",
       x = "Prediction Error (Actual minus Predicted)", y = "Density",
       fill = NULL, color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.subtitle = element_text(color = "grey40", size = 10))
```

The regression tree's error distribution has a distinctive multi-modal shape, with spikes at specific values. This is the bucket effect: because the tree only predicts 6 distinct values, errors cluster at multiples of the differences between those values. LASSO and linear regression produce smooth, bell-shaped error distributions centered near zero, which is the ideal behavior.

---

## Key Takeaways

**FICO score and loan term drive interest rate more than anything else.** The regression tree found this by using only those two variables out of 15. Linear regression and LASSO confirm it with their coefficient magnitudes. Lending Club's rate-setting logic is, at its core, a function of how creditworthy you are and how long you need the money.

**Linear models outperform the tree here.** The interest rate in this dataset increases relatively smoothly as FICO falls or loan term extends. Linear models are built for exactly this kind of smooth continuous relationship. The regression tree divides borrowers into buckets and assigns everyone in a bucket the same predicted rate, which is a crude approximation of what is actually a continuous function.

**LASSO and linear regression perform nearly identically.** This is common when the dataset is large, the signal is strong, and the features are not highly multicollinear. LASSO's main advantage shows up more clearly in smaller datasets with many noisy features. Here, best subset selection and LASSO converge on similar variable choices and similar errors.

**A 3.5% RMS error means the model's typical prediction is off by about 3.5 percentage points.** Given that rates range from 6% to 26%, this is a meaningful error, not a trivial one. Some of what drives individual rates cannot be captured by the features in this dataset. Lending Club also uses proprietary internal grades and loan history information that are not included here.
